---
title: "Comparison of ATERA and Xenium Samples and Their Enhancement with SPLIT"
output: 
  rmarkdown::github_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{ATERA vs Xenium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi     = 300,
  fig.width  = 7,
  fig.height = 4
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(Seurat)
library(arrow)
```

# Introduction

This vignette compares breast cancer samples profiled with ATERA and Xenium, then demonstrates the transcriptomic enhancement obtained with SPLIT, along with the new, more optimised and versatile SPLIT v0.2.0+.

# Overview

1. Data pre-processing
2. Side-by-side comparison of key metrics
3. Annotation with RCTD using a single-cell reference (pre-computed)
4. Evaluation of transcript spillover
5. Purification with SPLIT

## Data

We use publicly available matched breast cancer samples profiled with ATERA and Xenium from the same donor. For annotation, we use a publicly available annotated Chromium snRNA-seq breast cancer dataset (Janesick et al., 2023).

**Data sources:**

- ATERA: <https://www.10xgenomics.com/datasets/atera-wta-ffpe-human-breast-cancer>
- Xenium (sample S1 bottom): <https://www.10xgenomics.com/datasets/xenium-ffpe-human-breast-biomarkers>
- Chromium Flex (reference): <https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast>

```{r paths, include=FALSE}
atera_path  <- file.path("~/SPLIT/vignettes/data/atera")
xenium_path <- file.path("~/SPLIT/vignettes/data/xenium")
chrom_path  <- file.path("~/SPLIT/vignettes/data/chromium")
```

```{r load-data, include=FALSE}
# Pre-processed objects are loaded from disk; run load_10x_sample() / load_xenium_sample()
# (defined in aux.R) to regenerate them from raw data.
atera    <- readRDS("~/SPLIT/vignettes/data/atera.rds")
xenium   <- readRDS("~/SPLIT/vignettes/data/xenium.rds")
chromium <- readRDS("~/SPLIT/vignettes/data/chromium.rds")
```

# Side-by-side Comparison of Key Metrics

## QC

Xenium is profiled with a targeted panel of 269 genes, while ATERA provides full-transcriptome coverage (~18,000 genes). We compare distributions of counts per cell, detected genes per cell, and the counts-to-features ratio, first across each platform's full panel, then restricted to the shared gene set (n = 263).

```{r palette-exp, include=FALSE}
pal_exp <- c(
  "ATERA"    = "#E41A1C",
  "Xenium"   = "#377EB8",
  "Chromium" = "#1a1a1a"
)
```

### Full panel

```{r qc-full, fig.height=3.5}
qc_df <- bind_rows(
  atera@meta.data %>%
    select(nCount = nCount_Xenium, nFeature = nFeature_Xenium) %>%
    mutate(Experiment = "ATERA"),
  xenium@meta.data %>%
    select(nCount = nCount_Xenium, nFeature = nFeature_Xenium) %>%
    mutate(Experiment = "Xenium"),
  chromium@meta.data %>%
    select(nCount = nCount_RNA, nFeature = nFeature_RNA) %>%
    mutate(Experiment = "Chromium")
) %>%
  mutate(nCount_to_nFeature = nCount / nFeature) %>%
  pivot_longer(
    cols      = c(nCount, nFeature, nCount_to_nFeature),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  mutate(
    Metric     = ordered(metric, levels = c("nCount", "nFeature", "nCount_to_nFeature")),
    Experiment = ordered(Experiment, levels = c("Xenium", "ATERA", "Chromium"))
  )

ggplot(qc_df, aes(y = value, color = Experiment, x = Experiment)) +
  geom_boxplot(
    position     = position_dodge(width = 0.8),
    outlier.size = 0.15,
    width        = 0.65
  ) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    x     = NULL,
    y     = "Value",
    title = paste0("Xenium (", nrow(xenium), " genes) vs ATERA (",
                   nrow(atera), " genes) vs Chromium (", nrow(chromium), " genes)"),
    color = "Experiment"
  ) +
  scale_y_log10() +
  scale_color_manual(values = pal_exp) +
  theme_classic() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    aspect.ratio = 1,
    legend.position = "none"
  )
```

### Shared genes (n = 263)

```{r qc-common, fig.height=3.5}
common_genes <- intersect(rownames(chromium), intersect(rownames(atera), rownames(xenium)))

atera[["common"]]    <- CreateAssayObject(counts = GetAssayData(atera,    assay = "Xenium", layer = "counts")[common_genes, ])
xenium[["common"]]   <- CreateAssayObject(counts = GetAssayData(xenium,   assay = "Xenium", layer = "counts")[common_genes, ])
chromium[["common"]] <- CreateAssayObject(counts = GetAssayData(chromium, assay = "RNA",    layer = "counts")[common_genes, ])

qc_df_common <- bind_rows(
  atera@meta.data %>%
    select(nCount = nCount_common, nFeature = nFeature_common) %>%
    mutate(Experiment = "ATERA"),
  xenium@meta.data %>%
    select(nCount = nCount_common, nFeature = nFeature_common) %>%
    mutate(Experiment = "Xenium"),
  chromium@meta.data %>%
    select(nCount = nCount_common, nFeature = nFeature_common) %>%
    mutate(Experiment = "Chromium")
) %>%
  mutate(nCount_to_nFeature = nCount / nFeature) %>%
  pivot_longer(
    cols      = c(nCount, nFeature, nCount_to_nFeature),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  mutate(
    Metric     = ordered(metric, levels = c("nCount", "nFeature", "nCount_to_nFeature")),
    Experiment = ordered(Experiment, levels = c("Xenium", "ATERA", "Chromium"))
  )

ggplot(qc_df_common, aes(y = value, color = Experiment, x = Experiment)) +
  geom_boxplot(
    position     = position_dodge(width = 0.8),
    outlier.size = 0.15,
    width        = 0.65
  ) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    x     = NULL,
    y     = "Value",
    title = paste0("Xenium vs ATERA vs Chromium across ", length(common_genes), " shared genes"),
    color = "Experiment"
  ) +
  scale_y_log10() +
  scale_color_manual(values = pal_exp) +
  theme_classic() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    aspect.ratio = 1,
    legend.position = "none"
  )
```

**Summary:** ATERA provides full-transcriptome single-cell spatial profiling at a sensitivity comparable to Chromium.

---

## Cell-type Annotation with a Single-cell Reference

To annotate the matched ATERA and Xenium samples we used a previously published snRNA-seq breast cancer dataset (Janesick et al., 2023) as a reference. Annotation was performed with [rctd-py](https://github.com/p-gueguen/rctd-py), a fast Python reimplementation of [RCTD](https://github.com/dmcable/spacexr).

> **Note:** SPLIT v0.2.0+ is annotation-tool agnostic -- any deconvolution tool of your choice can be used (see the [SPLIT section](#split) below).

### Reference preparation

The original reference contains mixed-cell-type clusters ("T Cell & Tumor Hybrid", "Stromal & T Cell Hybrid") that are removed prior to annotation. Cell types are also consolidated into three levels of granularity (Level 1-3).

```{r reference-prep, message=FALSE, warning=FALSE}
chromium <- subset(
  chromium,
  subset = !cell_type %in% c("T Cell & Tumor Hybrid", "Stromal & T Cell Hybrid") & !is.na(cell_type)
)

chromium <- chromium %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA() %>%
  RunUMAP(dims = 1:50)
```

```{r level-maps, include=FALSE}
level2_map <- c(
  "B Cells"                  = "B cells",
  "CD4+ T Cells"             = "T cells",
  "CD8+ T Cells"             = "T cells",
  "Macrophages 1"            = "Myeloid",
  "Macrophages 2"            = "Myeloid",
  "IRF7+ DCs"                = "Myeloid",
  "LAMP3+ DCs"               = "Myeloid",
  "Mast Cells"               = "Myeloid",
  "DCIS 1"                   = "Pre-invasive",
  "DCIS 2"                   = "Pre-invasive",
  "Invasive Tumor"           = "Tumor",
  "Prolif Invasive Tumor"    = "Tumor",
  "Myoepi ACTA2+"            = "Myoepithelial",
  "Myoepi KRT15+"            = "Myoepithelial",
  "Stromal"                  = "Fibroblast",
  "Endothelial"              = "Endothelial / pericyte",
  "Perivascular-Like"        = "Endothelial / pericyte",
  "Stromal & T Cell Hybrid"  = "Hybrid",
  "T Cell & Tumor Hybrid"    = "Hybrid"
)

level1_map <- c(
  "B cells"                = "Immune",
  "T cells"                = "Immune",
  "Myeloid"                = "Immune",
  "Pre-invasive"           = "Epithelial",
  "Tumor"                  = "Epithelial",
  "Myoepithelial"          = "Epithelial",
  "Fibroblast"             = "Stromal / vascular",
  "Endothelial / pericyte" = "Stromal / vascular",
  "Hybrid"                 = "Hybrid / ambiguous"
)

chromium$Level3 <- unname(chromium$cell_type)
chromium$Level2 <- unname(level2_map[chromium$Level3])
chromium$Level1 <- unname(level1_map[chromium$Level2])
```

```{r palettes, include=FALSE}
pal_l3 <- ggsci::pal_igv()(length(unique(chromium$Level3)))
names(pal_l3) <- sort(unique(chromium$Level3))

pal_l2 <- ggsci::pal_igv()(length(unique(chromium$Level2)))
names(pal_l2) <- sort(unique(chromium$Level2))

pal_l1 <- ggsci::pal_igv()(length(unique(chromium$Level1)))
names(pal_l1) <- sort(unique(chromium$Level1))
```

```{r umap-reference, fig.height=6}
p_l3 <- UMAPPlot(chromium, group.by = "Level3", cols = pal_l3, label = T, repel = T) &
  theme_void() & theme(aspect.ratio = 1, legend.position = "none") & ggtitle("Level 3")

p_l2 <- UMAPPlot(chromium, group.by = "Level2", cols = pal_l2, label = T, repel = T) &
  theme_void() & theme(aspect.ratio = 1, legend.position = "none") & ggtitle("Level 2")

p_l1 <- UMAPPlot(chromium, group.by = "Level1", cols = pal_l1, label = T, repel = T) &
  theme_void() & theme(aspect.ratio = 1, legend.position = "none") & ggtitle("Level 1")

p_l3 | p_l2 | p_l1
```

### RCTD annotation results

RCTD was run at Level 2 using Level 1 as `class_df` to guide the method on related cell types.

```{r load-rctd, include=FALSE}
rctd_xenium <- readRDS("~/SPLIT/vignettes/data/rctd_xenium.rds")
rctd_atera  <- readRDS("~/SPLIT/vignettes/data/rctd_atera.rds")

xenium <- AddMetaData(xenium, rctd_xenium@results$results_df)
atera  <- AddMetaData(atera,  rctd_atera@results$results_df)

# Add spatial coordinates as a DimReduc object
sp_coords_xe            <- as.matrix(xenium@meta.data %>% select(x_centroid, y_centroid))
colnames(sp_coords_xe)  <- c("ST_1", "ST_2")
xenium[["spatial"]]     <- CreateDimReducObject(sp_coords_xe, assay = "Xenium", key = "ST_")

sp_coords_at            <- as.matrix(atera@meta.data %>% select(x_centroid, y_centroid))
colnames(sp_coords_at)  <- c("ST_1", "ST_2")
atera[["spatial"]]      <- CreateDimReducObject(sp_coords_at, assay = "Xenium", key = "ST_")
```

```{r spatial-annotation, fig.height=5}
p_xe_sp <- DimPlot(xenium, group.by = "first_type", label = F, cols = pal_l2,
                   reduction = "spatial", raster = T) +
  theme_void() + coord_fixed() + ggtitle("Xenium")

p_atera_sp <- DimPlot(atera, group.by = "first_type", label = F, cols = pal_l2,
                      reduction = "spatial", raster = T) +
  theme_void() + coord_fixed() + ggtitle("ATERA")

p_xe_sp + p_atera_sp +
  plot_layout(guides = "collect", ncol = 2) &
  theme(legend.position = "bottom", legend.direction = "horizontal") &
  guides(color = guide_legend(nrow = 1, override.aes = list(size = 3)))
```

```{r proportion-comparison, fig.height=4}
# Primary cell-type proportions
df_first_xe <- as.data.frame(table(xenium$first_type, useNA = "ifany")) %>%
  setNames(c("CellType", "Count")) %>%
  dplyr::arrange(desc(Count))

df_first_atera <- as.data.frame(table(atera$first_type, useNA = "ifany")) %>%
  setNames(c("CellType", "Count")) %>%
  dplyr::arrange(desc(Count))

plot_df <- df_first_xe %>%
  dplyr::rename(Count_first_xe = Count) %>%
  dplyr::inner_join(df_first_atera %>% dplyr::rename(Count_first_atera = Count), by = "CellType") %>%
  dplyr::mutate(
    Prop_first_xe    = Count_first_xe    / sum(Count_first_xe),
    Prop_first_atera = Count_first_atera / sum(Count_first_atera)
  )

p_primary_xe_atera <- ggplot(plot_df,
    aes(x = Prop_first_xe, y = Prop_first_atera, label = CellType, color = CellType)) +
  geom_point(size = 3) +
  geom_text_repel(size = 4, max.overlaps = Inf) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = pal_l2, na.value = "grey50") +
  labs(
    x     = "Primary proportion - Xenium",
    y     = "Primary proportion - ATERA",
    title = "Primary cell-type proportions"
  ) +
  theme_classic() +
  theme(legend.position = "none", aspect.ratio = 1)

# Secondary cell-type proportions
df_second_xe <- as.data.frame(table(xenium$second_type, useNA = "ifany")) %>%
  setNames(c("CellType", "Count")) %>%
  dplyr::mutate(CellType = as.character(CellType) %>% tidyr::replace_na("NA")) %>%
  dplyr::arrange(desc(Count))

df_second_atera <- as.data.frame(table(atera$second_type, useNA = "ifany")) %>%
  setNames(c("CellType", "Count")) %>%
  dplyr::mutate(CellType = as.character(CellType) %>% tidyr::replace_na("NA")) %>%
  dplyr::arrange(desc(Count))

plot_df_second <- df_second_xe %>%
  dplyr::rename(Count_second_xe = Count) %>%
  dplyr::full_join(df_second_atera %>% dplyr::rename(Count_second_atera = Count), by = "CellType") %>%
  dplyr::mutate(
    Count_second_xe    = tidyr::replace_na(Count_second_xe, 0),
    Count_second_atera = tidyr::replace_na(Count_second_atera, 0),
    Prop_second_xe     = Count_second_xe    / sum(Count_second_xe),
    Prop_second_atera  = Count_second_atera / sum(Count_second_atera)
  )

p_second_xe_atera <- ggplot(plot_df_second,
    aes(x = Prop_second_xe, y = Prop_second_atera, label = CellType, color = CellType)) +
  geom_point(size = 3) +
  geom_text_repel(size = 4, max.overlaps = Inf) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = pal_l2, na.value = "grey50") +
  labs(
    x     = "Secondary proportion - Xenium",
    y     = "Secondary proportion - ATERA",
    title = "Secondary cell-type proportions"
  ) +
  theme_classic() +
  theme(legend.position = "none", aspect.ratio = 1)

p_primary_xe_atera + p_second_xe_atera
```

### Transcript spillover

The secondary cell-type weight (`w2`) estimates the fraction of transcripts attributed to a contaminating neighbouring cell type, serving as a proxy for transcript spillover.

```{r spillover, fig.height=3.5}
plot_df_w2 <- bind_rows(
  atera@meta.data  %>% select(weight_second_type) %>% mutate(dataset = "ATERA"),
  xenium@meta.data %>% select(weight_second_type) %>% mutate(dataset = "Xenium")
)

median_df <- plot_df_w2 %>%
  group_by(dataset) %>%
  summarise(median_weight = median(weight_second_type, na.rm = TRUE), .groups = "drop")

ggplot(plot_df_w2, aes(x = weight_second_type, color = dataset, fill = dataset)) +
  geom_density(alpha = 0.2, linewidth = 1) +
  geom_vline(
    data     = median_df,
    aes(xintercept = median_weight, color = dataset),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  scale_fill_manual(values  = pal_exp) +
  scale_color_manual(values = pal_exp) +
  labs(
    x     = "w2 (contamination weight)",
    y     = "Density",
    title = "Transcript spillover: Xenium vs ATERA",
    color = "Experiment",
    fill  = "Experiment"
  ) +
  theme_classic() +
  theme(aspect.ratio = 1)
```

**Summary:** There is strong agreement between ATERA and Xenium annotations despite the former being derived from ~1,600 shared genes and the latter from 183, demonstrating that both technologies are compatible with label transfer. ATERA shows lower but detectable transcript spillover compared to Xenium. This reduced spillover is also evident in the UMAP embeddings below, where cell types in ATERA are better separated.

---

## UMAP embeddings (pre-SPLIT)

```{r umap-preprocessing, message=FALSE, warning=FALSE}
xenium <- xenium %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA() %>%
  RunUMAP(dims = 1:30)

atera <- atera %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA() %>%
  RunUMAP(dims = 1:50)
```

```{r umap-raw, fig.height=5}
p_umap_xe <- UMAPPlot(xenium, group.by = "first_type", label = T, repel = T,
                      cols = pal_l2, raster = F) +
  theme_void() + ggtitle("Raw Xenium") + NoLegend()

p_umap_atera <- UMAPPlot(atera, group.by = "first_type", label = T, repel = T,
                         cols = pal_l2, raster = F) +
  theme_void() + ggtitle("Raw ATERA") + NoLegend()

p_umap_xe | p_umap_atera
```

---

# SPLIT

SPLIT is a statistical decontamination method that removes transcript spillover from spatial transcriptomics data. It is compatible with any single-cell technology, including large-scale full-transcriptome platforms such as ATERA, and is annotation-tool agnostic as of v0.2.0.

```{r split-xenium, message=FALSE, warning=FALSE}
split_xenium <- SPLIT::purify(
  counts             = GetAssayData(xenium, assay = "Xenium", layer = "counts"),
  rctd               = rctd_xenium,
  DO_output_sce      = FALSE,
  chunk_size         = 10000,
  DO_purify_singlets = TRUE
)

purified_xenium <- CreateSeuratObject(
  counts    = split_xenium$purified_counts,
  meta.data = split_xenium$cell_meta,
  assay     = "Xenium"
)

purified_xenium <- subset(purified_xenium, subset = nCount_Xenium > 10)
purified_xenium <- purified_xenium[
  rowSums(GetAssayData(purified_xenium, assay = "Xenium", layer = "counts") > 0) >= 5, ]

purified_xenium <- purified_xenium %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>%
  RunPCA(verbose = FALSE) %>%
  RunUMAP(dims = 1:30, verbose = FALSE)
```

```{r split-atera, message=FALSE, warning=FALSE}
split_atera <- SPLIT::purify(
  counts             = GetAssayData(atera, assay = "Xenium", layer = "counts"),
  rctd               = rctd_atera,
  DO_output_sce      = FALSE,
  chunk_size         = 10000,
  DO_purify_singlets = TRUE
)

purified_atera <- CreateSeuratObject(
  counts    = split_atera$purified_counts,
  meta.data = split_atera$cell_meta,
  assay     = "Xenium"
)

purified_atera <- subset(purified_atera, subset = nCount_Xenium > 10)
purified_atera <- purified_atera[
  rowSums(GetAssayData(purified_atera, assay = "Xenium", layer = "counts") > 0) >= 5, ]

purified_atera <- purified_atera %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>%
  RunPCA(verbose = FALSE) %>%
  RunUMAP(dims = 1:50, verbose = FALSE)
```

```{r umap-post-split, fig.height=5}
p_umap_xe_split <- UMAPPlot(purified_xenium, group.by = "first_type", label = T, repel = T,
                             cols = pal_l2, raster = F) +
  theme_void() + ggtitle("Post-SPLIT Xenium") + NoLegend()

p_umap_atera_split <- UMAPPlot(purified_atera, group.by = "first_type", label = T, repel = T,
                                cols = pal_l2, raster = F) +
  theme_void() + ggtitle("Post-SPLIT ATERA") + NoLegend()

p_umap_xe_split | p_umap_atera_split
```

---
## Annotstion-tool-agnostic SPLIT
As of v0.2.0, SPLIT does not depend of RCTD output. To run SPLIT, you just need a weight matrix, a vector of primary cell-type and a reference profile. See below how to run SPLIT with any decomposition below:

```{r}

## Here we extract output of RCTD, but what we need is just a deconvolution weights, primary cell-type vector and the reference matrix

new_input <- SPLIT::convert_rctd_result_to_purify_input(rctd = rctd_xenium)

# Run SPLIT purification
split_xenium_v020 <- SPLIT::purify(
  counts = GetAssayData(xenium, assay = 'Xenium', layer = 'counts'), # or any gene x cells counts matrix
  reference = t(new_input$reference),
  primary_cell_type = new_input$primary_cell_type,
  deconvolution_weights = new_input$deconvolution_weights,
  DO_output_sce = TRUE
  )
```

---
# Session Info

```{r session-info}
sessionInfo()
```


